---
title: "Eigenfaces"
output: pdf_document
date: "2025-02-12"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(OpenImageR)
library(EBImage)
library(grid)
```

# Notes and Teach inputs

Assignment 1
raw_data -> KNN
raw_data -> PCA -> KNN

Steps:
1. Intra Class distances (same person) -> Visualize resulting histograms
2. Between Class distances (different persons) 
3. Set Threshold between Distance BetweeenClasses and Distance InterClass
4. Use difference Distances
5. Use PCA 

Assignment 2
1. FDA
raw_data -> PCA -> DiscrAnal(Fisher Faces) -> KNN

# Data Import and Preprocessing 

```{r}
# load selfmade functions
source("PCA_function.R")
source("read_data_function.R")
```


# Data Import and Preprocessing 

```{r}
folder_path = "Training"
data = read_all_images(folder_path)
```


# PCA

```{r, echo=FALSE, warning=FALSE, message=FALSE}

```

```{r}
### Verify with library
out = princomp(X)

pc1 = matrix(out$scores[,1],nrow=nrow(red), ncol=ncol(green))
imageShow(pc1)

pc2 = matrix(out$scores[,2],nrow=nrow(red), ncol=ncol(green))
imageShow(pc2)

pc3 = matrix(out$scores[,3],nrow=nrow(red), ncol=ncol(green))
imageShow(pc3)
```


## KNN

```{r}
# TODO: Histograms of distances
```


### Set Up Threshold Functions

```{r}
### Percentile of within distance
estimate_threshold_percentile <- function(train_data, train_labels, percentile = 0.90) {
  class_1 <- train_data[train_labels == 1, , drop = FALSE]
  dist_1 <- as.matrix(dist(class_1))
  
  # Extract upper triangle distances
  threshold <- quantile(dist_1[upper.tri(dist_1)], percentile)

  return(threshold)
}
```

### Set up Metric

```{r}
# TODO: Metrics
```

### Test/Train Split

```{r}
# TODO: Test/Train Split
```


### Var1: Groupwise distance thrashhold
```{r}

### KNN Groupwise Algo
knn_groupwise <- function(train_data, train_labels, test_data, k, threshold) {
  
  # get intergroup distances for each group
  for (i in 1:nrow(test_data)) {
    
    distances <- as.matrix(dist(rbind(test_data[i, ], train_data)))[1, -1]
    # Check treshhold for class 0
    avg_distance <- mean(distances[k_neighbors])
    
    predictions <- c(predictions, predicted_label)
  }
  
  # Check if distance threshold is passed for any group 
  if (avg_distance > threshold) {
    predicted_label <- "0"
  } else {
    # Run KNN for all groups where threshhol was passed  
    # TODO: Set K
    k_neighbors <- order(distances)[1:k]
    neighbor_labels <- train_labels[k_neighbors] # get labels of neighbors
    predicted_label <- names(sort(table(neighbor_labels), decreasing = TRUE))[1] # make frequency table and pick most frequent 
    predicted_label <- names(sort(table(neighbor_labels), decreasing = TRUE))[1]
  }
  
  return(predictions)
}
```


### Var2: Population Based
```{r}
### KNN Population Algo
knn_pop <- function(train_data, train_labels, test_data, k, threshold) {
  predictions <- c()
  
  ###
  for (i in 1:nrow(test_data)) {
    distances <- as.matrix(dist(rbind(test_data[i, ], train_data)))[1, -1]
    avg_distance <- mean(distances[k_neighbors])

    # Check if Population Threshold is passed
    if (avg_distance > threshold) {
      predicted_label <- "0"
    } else {
      # Run KNN for all groups
      # TODO: Optimize K
      k_neighbors <- order(distances)[1:k]
      neighbor_labels <- train_labels[k_neighbors] # get labels of neighbors
      predicted_label <- names(sort(table(neighbor_labels), decreasing = TRUE))[1] # make frequency table and pick most frequent 
      predicted_label <- names(sort(table(neighbor_labels), decreasing = TRUE))[1]
    }
    
    predictions <- c(predictions, predicted_label)
  }
  
  return(predictions)
}

```

### Run KNN
```{r}
# TODOS: 
# 1. Groupwise for with/without PCA, with different Threshholds
# 2. Population Based for with/without PCA, with different Threshholds


test_image <- load_and_flatten("Test_alpha/20BT.jpg")  # Load & flatten test image
test_data <- matrix(test_image, nrow = 1)  # Ensure test data format


# Compute threshold
threshold <- estimate_threshold_percentile(train_data, train_labels,percentile= 0.90)
print(threshold)

k <- 4  # Choose k value
prediction <- knn_manual(train_data, train_labels, test_data, k,threshold=threshold)

print(paste("Person_ID:", prediction))

```

### Compare Results

TODO: Discuss Results

```{r}
# TODO: Display Results
```


# LDS/FDS

```{r}
# TODO: Linear Discriminant Analysis/Fisher discriminant Analysis
```













