---
title: "Eigenfaces"
output: pdf_document
date: "2025-02-12"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(OpenImageR)
library(EBImage)
library(grid)
library(ggplot2)

```

# Notes and Teach inputs

Assignment 1
raw_data -> KNN
raw_data -> PCA -> KNN

Steps:
1. Intra Class distances (same person) -> Visualize resulting histograms
2. Between Class distances (different persons) 
3. Set Threshold between Distance BetweeenClasses and Distance InterClass
4. Use difference Distances
5. Use PCA 

Assignment 2
1. FDA
raw_data -> PCA -> DiscrAnal(Fisher Faces) -> KNN

# Data Import and Preprocessing 

```{r}
# load selfmade functions
source("PCA_function.R")
source("read_data_function.R")
```


# Data Import and Preprocessing 

```{r}
folder_path = "Training"
data = read_all_images(folder_path)
```


# PCA

```{r, echo=FALSE, warning=FALSE, message=FALSE}

```

```{r}
### Verify with library
out = princomp(X)

pc1 = matrix(out$scores[,1],nrow=nrow(red), ncol=ncol(green))
imageShow(pc1)

pc2 = matrix(out$scores[,2],nrow=nrow(red), ncol=ncol(green))
imageShow(pc2)

pc3 = matrix(out$scores[,3],nrow=nrow(red), ncol=ncol(green))
imageShow(pc3)
```


## KNN

### Make Matrix

```{r}
data_matriz = as.matrix(data)
```


### Set Up Data


```{r}

distance_distributions <- function(data, distance_metric="euclidean"){
  
  ids <- data[, ncol(my_matrix)]
  
  # set up distance matrix/data frame
  dist_matrix <- as.matrix(dist(data[, -ncol(data)], method = distance_metric))
  dist_df <- data.frame(
    row1 = rep(1:nrow(data), each = nrow(data)),
    row2 = rep(1:nrow(data), times = nrow(data)),
    distance = as.vector(dist_matrix)
  )
  dist_df <- dist_df[dist_df$row1 != dist_df$row2, ] # remove self distances
  
  dist_df$id1 <- ids[dist_df$row1] # Assign IDs
  dist_df$id2 <- ids[dist_df$row2]
  
  dist_df_same_id <- dist_df[dist_df$id1 == dist_df$id2, ] # In group distances
  dist_df_same_id <- dist_df_same_id[order(dist_df_same_id$id1), ]
  
  
  dist_df_diff_id <- dist_df[dist_df$id1 != dist_df$id2, ] # Outside group distances
  
  dist_df_diff_id$id1 = 0
  dist_df_diff_id$id2 = 0
  
  dist_df_split <- rbind(dist_df_same_id,dist_df_diff_id)
  
  
  # Plot histogram
  plot = ggplot(dist_df_split, aes(x = distance, fill = factor(id1))) +
    geom_density(alpha = 0.6, position = "identity") +
    labs(title = "Histogram of Pairwise Distances by ID",
         x = "Distance", y = "Frequency", fill = "ID Group") +
    theme_minimal()
  
  return(plot)
    
}

```

```{r}
# Run
for (distance_metric in c("euclidean", "manhattan")) {
  hist = distance_distributions(data_matriz, distance_metric = distance_metric)
  print(hist)
}

```

### Set Up Threshold Functions

```{r}
### Percentile of within distance
estimate_threshold_percentile <- function(train_data, train_labels, percentile = 0.90) {
  class_1 <- train_data[train_labels == 1, , drop = FALSE]
  dist_1 <- as.matrix(dist(class_1))
  
  # Extract upper triangle distances
  threshold <- quantile(dist_1[upper.tri(dist_1)], percentile)

  return(threshold)
}
```

### Set up Metric

```{r}
# TODO: Metrics
score <- function(true_labels, predicted_labels) {
  # Binary Classification Accuracy (In or not In DB)
  
  # TODO: ROC Curve
  # TODO: True Pos Rate
  # TODO: False Negative rate
  
  # Classification of Face ID
  mean(true_labels == predicted_labels)
  
  return()
}
```

### Test/Train Split

```{r}
# TODO: Test/Train Split

train_test_split <- function(data, train_ratio = 0.8, seed = NULL) {
  if (!is.null(seed)) {
    set.seed(seed)
  }
  # Split
  n = nrow(data)
  train_indices <- sample(1:n, size = floor(train_ratio * n))
  train_data <- data[train_indices, , drop = FALSE]  # Train set
  test_data <- data[-train_indices, , drop = FALSE]  # Test set
  
  return(list(train = train_data, test = test_data))
}
```


```{r}
split_data <- train_test_split(data = data_matriz, train_ratio = 0.8, seed = NULL)
train_data <- split_data$train
test_data <- split_data$test
```


### Var1: Groupwise distance thrashhold



```{r}
### Set up Thresholds

train_data_ids <- unique(train_data[, ncol(train_data)])

# precentile threshold function
estimate_threshold_via_percentile <- function(data, train_labels, percentile = 0.95) {
  dist_1 <- as.matrix(dist(data))
  threshold <- quantile(dist_1[upper.tri(dist_1)], percentile)
  return(threshold)
}

treshholds = list()
for (id in train_data_ids){ # for each id
  train_data_grouped <- train_data[train_data[, ncol(train_data)] == id, , drop = FALSE]
  treshholds[id] <- estimate_threshold_via_percentile(train_data_grouped) 
}

```


```{r}
# generate distances between test and train data
distance_list = list()
for (i in 1:nrow(test_data)) {
  test_row <- test_data[i, -ncol(test_data), drop = FALSE]
  train_data_no_last_col <- train_data[, -ncol(train_data), drop = FALSE]
  combined_data <- rbind(test_row, train_data_no_last_col)
  
  distance_matrix <- as.matrix(dist(combined_data)) #### EXTREMLY INEFFICIENT
  distances <- distance_matrix[1, -1] 
  
  # Store distances using test_data ID as key
  test_id <- test_data[1, ncol(test_data)]
  #names(distances) <- train_data$id  # Assign names to match training IDs
  distance_list[[as.character(test_id)]] <- distances

}
```


```{r}
# check if in group and if so, run knn
k = 10
for (id in as.numeric(names( ))){
  print(id)
  id_string = paste(strsplit(as.character(id), NULL)[[1]], collapse = "")
  print(id_string)
  if (any(distances < treshholds[id_string])){
      k_neighbors <- order(distances)[1:k]
      neighbor_labels <- train_data_ids[k_neighbors] # get labels of neighbors
      predicted_label <- names(sort(table(neighbor_labels), decreasing = TRUE))[1] # make frequency table and pick most frequent 
      predicted_label <- names(sort(table(neighbor_labels), decreasing = TRUE))[1]
      break
  }
  else{
    predicted_label <- "0"
  }
}
```

```{r}
avg_distance <- mean(distances[k_neighbors])
predictions <- c(predictions, predicted_label)

}

# Check if distance threshold is passed for any group 
if (avg_distance > threshold) {
  predicted_label <- "0"
} else {
  # Run KNN for all groups where threshhol was passed  
  # TODO: Set K

}
```


```{r}
for (id in train_data_ids){ # for each id
  train_data_id <- train_data[train_data[, ncol(train_data)] == id, , drop = FALSE]
  
  # get average distances within id group
  pairwise_distances <- as.vector(dist(train_data_id[, -ncol(train_data_id)]))
  avg_distance <- mean(distances[k_neighbors])

  for (i in 1:nrow(train_data_id)){
    distances <- as.matrix(dist(rbind(test_data[i, ], train_data_id)))[1, -1]
    
    predictions <- c(predictions, predicted_label)
  }
  
  # Check if distance threshold is passed for any group 
  if (avg_distance > threshold) {
    predicted_label <- "0"
  } else {
    # Run KNN for all groups where threshhol was passed  
    # TODO: Set K
    k_neighbors <- order(distances)[1:k]
    neighbor_labels <- train_labels[k_neighbors] # get labels of neighbors
    predicted_label <- names(sort(table(neighbor_labels), decreasing = TRUE))[1] # make frequency table and pick most frequent 
    predicted_label <- names(sort(table(neighbor_labels), decreasing = TRUE))[1]
  }
  
}



```


```{r}

### KNN Groupwise Algo
knn_groupwise <- function(data_train, data_test, k, threshold) {
  
  # get intergroup distances for each group
  for (i in 1:nrow(test_data)) {
    
    distances <- as.matrix(dist(rbind(test_data[i, ], data_train)))[1, -1]
    avg_distance <- mean(distances[k_neighbors])
    
    predictions <- c(predictions, predicted_label)
  }
  
  # Check if distance threshold is passed for any group 
  if (avg_distance > threshold) {
    predicted_label <- "0"
  } else {
    # Run KNN for all groups where threshhol was passed  
    # TODO: Set K
    k_neighbors <- order(distances)[1:k]
    neighbor_labels <- train_labels[k_neighbors] # get labels of neighbors
    predicted_label <- names(sort(table(neighbor_labels), decreasing = TRUE))[1] # make frequency table and pick most frequent 
    predicted_label <- names(sort(table(neighbor_labels), decreasing = TRUE))[1]
  }
  
  return(predictions)
}

```





```{r}

# Compute threshold
threshold <- estimate_threshold_percentile(train_data, train_labels,percentile= 0.90)
print(threshold)

k <- 4  # Choose k value
prediction <- knn_manual(train_data, train_labels, test_data, k,threshold=threshold)

print(paste("Person_ID:", prediction))
```



### Var2: Population Based
```{r}
### KNN Population Algo
knn_pop <- function(train_data, train_labels, test_data, k, threshold) {
  predictions <- c()
  
  ###
  for (i in 1:nrow(test_data)) {
    distances <- as.matrix(dist(rbind(test_data[i, ], train_data)))[1, -1]
    avg_distance <- mean(distances[k_neighbors])

    # Check if Population Threshold is passed
    if (avg_distance > threshold) {
      predicted_label <- "0"
    } else {
      # Run KNN for all groups
      # TODO: Optimize K
      k_neighbors <- order(distances)[1:k]
      neighbor_labels <- train_labels[k_neighbors] # get labels of neighbors
      predicted_label <- names(sort(table(neighbor_labels), decreasing = TRUE))[1] # make frequency table and pick most frequent 
      predicted_label <- names(sort(table(neighbor_labels), decreasing = TRUE))[1]
    }
    
    predictions <- c(predictions, predicted_label)
  }
  
  return(predictions)
}

```

### Run best KNN
```{r}

```

### Compare Results

TODO: Discuss Results

```{r}
# TODO: Display Results
```


# LDS/FDS

```{r}
# TODO: Linear Discriminant Analysis/Fisher discriminant Analysis
```













